---
title: "5: Part 1 - Data Visualization Basics"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## Objectives
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## Opening discussion
Effective data visualization depends on purposeful choices about graph types. The ideal graph type depends on the type of data and the message the visualizer desires to communicate. The best visualizations are clear and simple. A good resource for data visualization is [Data to Viz](https://www.data-to-viz.com/), which includes both a decision tree for visualization types and explanation pages for each type of data, including links to R resources to create them. Take a few minutes to explore this website.

## Set Up

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)
install.packages("ggridges")
library(ggridges) #for visualisation of variables over time

PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
PeterPaul.chem.nutrients.gathered <-
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv", stringsAsFactors = TRUE)
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv", stringsAsFactors = TRUE)

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
PeterPaul.chem.nutrients.gathered$sampledate <- as.Date(PeterPaul.chem.nutrients.gathered$sampledate, format = "%Y-%m-%d")

```

## ggplot

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_density_ridges
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot and Line Plot

```{r, fig.height = 3, fig.width = 4, warning = FALSE} 
#line above defines dimensions for all plots
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + #ggplot (dataframe, aesthetics (x,y))
  geom_point() #for a point

O3plot <- ggplot(EPAair) + #create object to assign ggplots/layers to print plot
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)

# Fix this code
O3plot2 <- ggplot(EPAair) +
  #geom_point(aes(x = Date, y = Ozone, color = "blue")) #when specifying within aesthetics, adding layer to plot, not changing colour of plot. colour that comes out is whatever comes first on ggplot
  geom_point(aes(x = Date, y = Ozone), color = "blue")
print(O3plot2)

# Add additional variables
# How could you automatically assign a marker color to a variable? 
PMplot <- 
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year), color = Site.Name)) + #automatically assign a shape based on year, colour is function of site name
  geom_point()
print(PMplot)

# Separate plot with facets
PMplot.faceted <-
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3) #create 1 plot for each site, instead of colouring each site by diff colours
print(PMplot.faceted)

# Filter dataset within plot building and facet by multiple variables
PMplot.faceted2 <- 
  ggplot(subset(EPAair, Site.Name == "Clemmons Middle" | Site.Name == "Leggett" |
                  Site.Name == "Bryson City"), 
         aes(x = Month, y = PM2.5)) + 
  geom_point() +
  facet_grid(Site.Name ~ Year)  #column represent year and row representing the sites
print(PMplot.faceted2)

# Plot true time series with geom_line
PMplot.line <- 
  ggplot(subset(EPAair, Site.Name == "Leggett"), 
         aes(x = Date, y = PM2.5)) +
  geom_line()
print(PMplot.line)


```


### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20)
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
tempvsdepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() #depth is reversed scale for oceans
print(tempvsdepth)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) + #coloured by depth. surface darker blue then move to lighter blue as it goes deeper into lake
  geom_point() +
  geom_smooth(method = lm) + #lm: linear model. add trendline
  geom_abline(aes(slope = 16, intercept = 0)) #add your own line (black line)
print(NvsP)

```

### Plotting continuous vs. categorical variables
A traditional way to display summary statistics of continuous variables is a bar plot with error bars. Let's explore why this might not be the most effective way to display this type of data. Navigate to the Caveats page on Data to Viz (https://www.data-to-viz.com/caveats.html) and find the page that explores barplots and error bars. 

What might be more effective ways to display the information? Navigate to the boxplots page in the Caveats section to explore further.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"? colour regards to contour of shape. fill is the fill of shape. if use fill you just change the colour of the boxplot but not the outline. use colour for boxplot
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient, fill = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge", dotsize = 2) #binaxis=y stacking dots on y
print(Nutrientplot4) #position=dodge -> dont overlay dots that fall on the same position

# Violin plot, density plot
Nutrientplot5 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "darkred") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "blue") +
  geom_freqpoly(aes(x = no23), color = "royalblue") +
  geom_freqpoly(aes(x = po4), color = "red") #no auto legends for this entire plot
print(Nutrientplot6)

# Using a gathered dataset, easier way to show ^
Nutrientplot7 <-   
  ggplot(PeterPaul.chem.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Frequency polygons have the risk of becoming spaghetti plots. 
# See <https://www.data-to-viz.com/caveat/spaghetti.html> for more info.

# Ridgeline plot
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(y = nutrient, x = concentration)) +
  geom_density_ridges(aes(fill = lakename), alpha = 0.5) #fill density ridges by lake (fill below the curve, not just the line of the graph)
print(Nutrientplot6)


```
