---
title: "6: Part 3 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe t-test under GLM framework

## Set up your session

```{r, message = FALSE}
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)

library(tidyverse)
library(lubridate)

```

## T-Test

Continuous response, one categorical explanatory variable with two categories (or comparison to a single value if a one-sample test).
h0: mean=b0
h1: u does not equal b0

### Formulating Hypothesis for µ

Two hypotheses are formed – the null hypothesis and the alternative hypothesis.
The null hypothesis and the alternative hypothesis combine to cover all possible values for the population mean.
The null hypothesis must have the equality.
The null and alternative hypotheses are always stated in terms of the population mean (mu).

### One-sample t-test
The object of a one sample test is to test the null hypothesis that the mean of the group is equal to a specific value. 

Function t.test()
**x**	a (non-empty) numeric vector of data values.
**alternative** a character string specifying the alternative hypothesis, must be one of "two.sided" (default), "greater" or "less". You can specify just the initial letter.
**mu** a number indicating the true value of the mean (or difference in means if you are performing a two sample test).
**formula** a formula of the form lhs ~ rhs where lhs is a numeric variable giving the data values and rhs either 1 for a one-sample or paired test or a factor with two levels giving the corresponding groups. If lhs is of class "Pair" and rhs is 1, a paired test is done.

The one-sample t-test relies on the assumption that the variable is normally distributed in the population. However, the t-test is robust to mild departures from normality when the sample size is small, and when the sample size is large the normality assumption becomes less important.


For example, we might ask ourselves (from the EPA air quality processed dataset): 
Are Ozone levels below the threshold for "good" AQI index (0-50)?

### Import data set

```{r, message = FALSE}
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv", stringsAsFactors = TRUE)
# Set date to date format
EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")

```


> Exercise 1: State the hypotheses for testing mean of AQI index.

> Answer: 
H0: mu >= 50
Ha: mu < 50

```{r}
summary(EPAair$Ozone) #mean = 40.88 and median = 40. 
length(EPAair$Ozone) #8976 observations. have alot of observations. mean of sample is good representation of the population. mean should be less than 50

O3.onesample <- t.test(EPAair$Ozone, mu = 50, alternative = "less") #alternative is less. cos of the hypothesis stated early. 
#if mu<=50, alternative = greater
#if mu=50, alternative is 2 sided. alternative is mu different than 50 
O3.onesample
#p value is <0.05, reject null, accept alternative. ie mean <50 

Ozone.plot <- ggplot(EPAair, aes(x = Ozone)) +
  geom_density(fill = "gray") + #density plot for variables
  geom_vline(xintercept = 50, color = "#238b45", lty = 2, size = 0.9) +
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0))
print(Ozone.plot)
#observations fall to left verticle line, which is in accordance to hypothesis

# Format as a GLM
O3.onesample2 <- lm(Ozone ~ 1, EPAair) #1 corresponds to intercept term. get mean of series for intercept value. 
O3.onesample2 #intercept term is the mean we had before
```

Write a sentence or two about the results of this test. Include both the results of the test and an interpretation that puts the findings in context of the research question.

> 

### Two-sample t-test

The two-sample *t* test is used to test the hypothesis that the mean of two samples is equivalent. Unlike the one-sample tests, a two-sample test requires a second assumption that the variance of the two groups is equivalent. 
H0: means are equal
h1: means are not equal. 

For example, we might ask ourselves (from the EPA air quality processed dataset): 
Are Ozone levels different between 2018 and 2019?

```{r}
#First let's look at the data
ggplot(EPAair, aes(x = Ozone, color = as.factor(Year))) +
  geom_freqpoly()
#hard to tell if mean is the same for both plots


#Format as a t-test
#EPAair$Ozone will be our continuous dependent variable
#EPAair$Year will be our categorical variable with two levels (2018 and 2019)
O3.twosample <- t.test(EPAair$Ozone ~ EPAair$Year)
O3.twosample
#p <0.05, reject null, accept alternative that 2 means are not equal. alternative hypothesis: true difference in means between group 2018 and group 2019 is not equal to 0


# Format as a GLM
O3.twosample2 <- lm(EPAair$Ozone ~ EPAair$Year) #ask if year is predictor of ozone level ? 
summary(O3.twosample2)
#for EPAair$year: p value <0.05, year is statistically different from 0 and is a predictor of ozone level

```





