---
title: 'Assignment 2: Coding Basics'
author: "Li Jia Go"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r generating a sequence}

#1. using sequence function to generate sequence of numbers from 1 to 100
# increasing by fours. i.e. seq (from, to, by)
seq(1,100,4) 

# generating sequence from 1 to 100, increasing by 4. naming sequence
# 'hundredseq'
hundredseq <- seq(1,100,4)

# return result
hundredseq 

#2. 
# computing the mean of the sequence using the mean function and assigning a
# name to the result
hundredseq_mean <- mean(hundredseq)

# return result
hundredseq_mean 

# computing the median of the sequence using the median function and assigning a
# name to the result
hundredseq_median <- median(hundredseq) 
hundredseq_median # return result


#3. 
hundredseq_mean > hundredseq_median # asking R if mean is greater than median
hundredseq_mean < hundredseq_median # asking R if mean is less than median
hundredseq_mean == hundredseq_median # asking R if mean and median are equal
hundredseq_mean != hundredseq_median # asking R if mean and median are not equal

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r creating vectors}
#5. 
#creating a vector of student names
student_names <- c ("John", "Peter", "Sarah", "Jill") 
student_names #return result

#creating a vector of student's scores
student_scores <- c(80, 90, 40, 48)
student_scores #return result

#creating a vector showing whether students have passed the test or not
student_pass <- c(TRUE, TRUE, FALSE, FALSE) 
student_pass #return result

#6. 
class(student_names) #character

class(student_scores) #numeric

class(student_pass) #logical

#7.
# transforming each vector into a data frame
df_classnames <- as.data.frame(student_names) 
df_classscores <- as.data.frame(student_scores)
df_classpass <- as.data.frame(student_pass)

# combining all the columns of the data frame
df_classtestscores <- cbind(df_classnames, df_classscores, df_classpass)
df_classtestscores #print result


#8.
# renaming names of columns of data frame
colnames(df_classtestscores) <- c("names", "scores", "passed") 
df_classtestscores #return result



```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Matrices can only store a single data type, while this data frame consists of different classes of data (i.e. characters, numbers, logic etc)

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r creating a function to see if given test score is a passing grade of 50 or above}

#creating a function to see if given test score is of a passing grade or not
passfifty<- function(x) { 
ifelse(x >50, "TRUE", "FALSE")
}

answer<-passfifty(student_scores)
print(answer) #print result


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Only 'ifelse' worked. The if() statement can only work with elements that are length 1, i.e. a single student. However, by using the 'ifelse' expression, we are able to check if the entire class of students passed the test at once.
