---
title: "Assignment 09: Data Scraping"
author: "Li Jia Go"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```
## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1

#checking working directory
getwd()

#loading relevant packages
library(tidyverse)
library(rvest)
library(dplyr)

#setting ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top", plot.title=element_text(size=12))
theme_set(mytheme)



```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
DEQ_website <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022")


```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water.system.name <- DEQ_website %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()

pswid <- DEQ_website %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text
  
ownership <- DEQ_website %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text
  
max.withdrawals.mgd <- DEQ_website %>% html_nodes ('th~ td+ td') %>% html_text

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4

#creating dataframe and month in order of scrape
#rearranging month in chronological order
df_withdrawals <- data.frame("Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11,4,8,12),
                             "Year" = rep(2022,12),
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>%
                    mutate(Date=lubridate::my(paste(Month, "-", Year))) %>%
                    dplyr::arrange(Date) 


#5 line plot of maximum daily withdrawals across months for 2021
ggplot(df_withdrawals, aes(x=Date, y=Max_Withdrawals_mgd)) +
  geom_line(aes()) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + 
  ylab("Maximum Withdrawals (mgd)") +
  xlab("2022") +
  ggtitle("2022 maximum daily water withdrawals per month, Durham (NC)")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.

#creating function
scrape.it <- function(the_year, the_PWSID){
  
  #Retrieving the website contents 
  theDEQsite <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', the_PWSID, '&year=', the_year))
  
  #locating elements and reading text attributes into variables
  water.system.name <- theDEQsite %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
  pswid <- theDEQsite %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text
  ownership <- theDEQsite %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text
  max.withdrawals.mgd <- theDEQsite %>% html_nodes ('th~ td+ td') %>% html_text
  
  #creating the dataframe
 df_withdrawals_func <- data.frame("Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4,8,12),
                             "Year" = rep(the_year,12),
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>%
                    mutate(WaterSystemName= !!water.system.name,
                           PSWID=!!pswid,
                           Ownership=!!ownership,
                      Date=lubridate::my(paste(Month, "-", Year))) %>%
                    dplyr::arrange(Date)
 

  #Return the dataframe
  return(df_withdrawals_func)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
#scraping for Durham, 2015
Durham_2015 <-scrape.it(2015, "03-32-010")

#plotting max daily withdrawals for Durham, 2015
ggplot(Durham_2015, aes(x=Date, y=Max_Withdrawals_mgd)) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + 
  ylab("Maximum Withdrawals (mgd)") +
  xlab("Year") +
  ggtitle("2015 maximum daily water withdrawals per month, Durham (NC)")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, tidy=TRUE, tidy.opts=list(width.cutoff=40), message=FALSE}
#8
Asheville_2015 <-scrape.it(2015, "01-11-010")

#merging both datasets into one dataframe
AshevilleDurham_2015 <- rbind (Asheville_2015, Durham_2015)

#plotting Asheville and Durham's water withdrawals
ggplot(AshevilleDurham_2015, aes(x=Date, y=Max_Withdrawals_mgd)) +
  geom_line(aes(color=WaterSystemName)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + 
  labs (x="2015", y="Maximum Withdrawals (mgd)", color="State") +
  ggtitle("2015 maximum daily water withdrawals per month, Asheville & Durham (NC)")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9

decade <- rep(2010:2021)
AshevillePWSID <- '01-11-010'

df.Asheville.decade <- map(decade, scrape.it, the_PWSID=AshevillePWSID)

df.Asheville.decade <- bind_rows(df.Asheville.decade)

#plotting Asheville's water withdrawals, 2010-2019
ggplot(df.Asheville.decade, aes(x=Date, y=Max_Withdrawals_mgd)) +
  geom_line() +
  geom_smooth(method = 'loess', se= FALSE) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  ylab("Maximum Withdrawals (mgd)") +
  xlab("Year") +
  ggtitle("Water withdrawals, Asheville (NC), 2010-2019")

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
>Answer: The usage of water in Asheville was relatively stable from 2010-2018, with lower usage from 2013-2014. From 2019 onwards, there was a significant increase in usage of water.